---
ta---
title: "week02_exercises"
author: "David Mendez"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 2H1.

Suppose there are two species of panda bear. Both are equally common in the wild and live in the same places. They look exactly alike and eat the same food, and there is yet no genetic assay capable of telling them apart. They differ however in their family sizes. Species A gives birth to twins 10% of the time, otherwise birthing a single infant. Species B births twins 20% of the time, otherwise birthing singleton infants. Assume these numbers are known with certainty, from many years of field research. Now suppose you are managing a captive panda breeding program. You have a new female panda of unknown species, and she has just given birth to twins. What is the probability that her next birth will also be twins?

```{r}
Pr_pb1=0.5
Pr_pb2=0.5
Pr_twins_pb1=0.1
Pr_twins_pb2=0.2

Pr_twins=Pr_twins_pb1*Pr_pb1+Pr_twins_pb2*Pr_pb2

Pr_twins_twins = ((Pr_twins_pb1^2+Pr_twins_pb2^2)/2)/Pr_twins

print(Pr_twins_twins)
```

## 2H2.

Recall all the facts from the problem above. Now compute the probability that the panda we have is from species A, assuming we have observed only the first birth and that it was twins.

``` r
```

```{r}

Pr_pb1_twins = Pr_twins_pb1*Pr_pb1/Pr_twins
print(Pr_pb1_twins)
```

``` r
```

## 2H3.

Continuing on from the previous problem, suppose the same panda mother has a second birth and that it is not twins, but a singleton infant. Compute the posterior probability that this panda is species A.

```{r}
Pr_a=Pr_pb1_twins
Pr_singleton_a=1-Pr_twins_pb1
Pr_singleton = Pr_singleton_a*Pr_a + (1-Pr_twins_pb2)*(1-Pr_a)

Pr_a_singleton=Pr_singleton_a*Pr_a/Pr_singleton

print(Pr_a_singleton)
```

## 2H4.

A common boast of Bayesian statisticians is that Bayesian inference makes it easy to use all of the data, even if the data are of different types. So suppose now that a veterinarian comes along who has a new genetic test that she claims can identify the species of our mother panda. But the test, like all tests, is imperfect. This is the information you have about the test:

-   The probability it correctly identifies a species A panda is 0.8.

-   The probability it correctly identifies a species B panda is 0.65.

The vet administers the test to your panda and tells you that the test is positive for species A. First ignore your previous information from the births and compute the posterior probability that your panda is species A. Then redo your calculation, now using the birth data as well.

```{r}


```

# Chapter 3

## Intro

```{r}
library(tidyverse)
p_grid <- seq( from=0 , to=1 , length.out=1000 )
prior <- rep( 1 , 1000 )
likelihood <- dbinom( 6 , size=9 , prob=p_grid )
posterior <- likelihood * prior
posterior <- posterior / sum(posterior)
set.seed(100)
samples <- sample( p_grid , prob=posterior , size=1e4 , replace=TRUE )
```

## 3E1. 

How much posterior probability lies below p = 0.2?

```{r}
sum(posterior[p_grid<0.2])
```

## 3E2. 

How much posterior probability lies above p = 0.8?

```{r}
sum(posterior[p_grid>0.8])
```

## 3E3. 

How much posterior probability lies between p = 0.2 and p = 0.8?

```{r}
sum(posterior[p_grid>0.2 & p_grid<0.8])
```

## 3E4. 

20% of the posterior probability lies below which value of p?

```{r}
p=0.5166
sum(posterior[p_grid<p])
```

## 3E5. 

20% of the posterior probability lies above which value of p?

```{r}
p=0.76
sum(posterior[p_grid>p])
```

## 3E6. 

Which values of p contain the narrowest interval equal to 66% of the posterior probability?

```{r}

i_s <- c()
j_s <- c()
for (i in 1:1000){
  for (j in 1:1000){
    if (i<j){
      p_post=sum(posterior[i:j])
      if (round(p_post,digits=4)==0.66){
        i_s <- c(i_s,i)
        j_s <- c(j_s,j)
      }
    }
  }
}

data.frame(p_low=p_grid[i_s],
           p_high=p_grid[j_s]) %>% 
  mutate(inter=p_high-p_low) %>% 
  arrange(inter) %>% 
  head()
```

## 3E7. 

Which values of p contain 66% of the posterior probability, assuming equal posterior probability both below and above the interval?

```{r}
i_s <- c()
inter <- c()
for (i in 1:500){
  inter=c(inter,sum(posterior[i:(1001-i)]))
  i_s <- c(i_s,i)
}
data.frame(p_low=p_grid[i_s],
           p_high=p_grid[1001-i_s],
           inter=inter) %>% 
  filter(inter>0.659, inter<0.66)
```

## 3M1. 

Suppose the globe tossing data had turned out to be 8 water in 15 tosses. Construct the posterior distribution, using grid approximation. Use the same flat prior as before.

```{r}
p_grid <- seq( from=0 , to=1 , length.out=1000 )
prior <- rep( 1 , 1000 )
likelihood <- dbinom( 8 , size=15 , prob=p_grid )
posterior <- likelihood * prior
posterior <- posterior / sum(posterior)
```

## 3M2. 

Draw 10,000 samples from the grid approximation from above. Then use the samples to calculate the 90% HPDI for p.

```{r}
set.seed(31)
samples <- sample( p_grid , size=1e4 , replace=TRUE , prob=posterior )
rethinking::HPDI( samples , prob=0.5 )
```

## 3M3. 

Construct a posterior predictive check for this model and data. This means simulate the distribution of samples, averaging over the posterior uncertainty in p. What is the probability of observing 8 water in 15 tosses?

```{r}
w <- rbinom( 1e4 , size=15 , prob=samples )
sum(w==8)/1e4
simplehist(w)
```

## 3M4. 

Using the posterior distribution constructed from the new (8/15) data, now calculate the probability of observing 6 water in 9 tosses.

```{r}
sum(rbinom( 1e4 , size=9 , prob=samples )==6)/1e4
```

## 3M5. 

Start over at 3M1, but now use a prior that is zero below p = 0.5 and a constant above p = 0.5. This corresponds to prior information that a majority of the Earth’s surface is water. Repeat each problem above and compare the inferences. What difference does the better prior make? If it helps, compare inferences (using both priors) to the true value p = 0.7.

```{r}
prior_2 <- c(rep( 0 , 500 ),rep(5,500))
likelihood_2 <- dbinom( 8 , size=15 , prob=p_grid)
posterior_2 <- likelihood_2 * prior_2
posterior_2 <- posterior_2 / sum(posterior_2)
set.seed(100)
plot(p_grid, posterior_2,type="l")
```

## 3M6. 

Suppose you want to estimate the Earth’s proportion of water very precisely. Specifically, you want the 99% percentile interval of the posterior distribution of p to be only 0.05 wide. This means the distance between the upper and lower bound of the interval should be 0.05. How many times will you have to toss the globe to do this?

```{r}
inter <- c()
i_s <- c()
for (i in 1:100){
  samples_3 <- sample( c(T,F) , size=i , replace=TRUE , prob=c(0.3,0.7) )
  likelihood_3 <- dbinom( sum(samples_3) , size=i , prob=p_grid)
  posterior_3 <- likelihood_3 * prior_2
  posterior_3 <- posterior_3 / sum(posterior_3)
  samples_3 <- sample( p_grid , size=1e4 , replace=TRUE , prob=posterior_3 )
  inter <- c(inter, HPDI(samples,prob=0.99)[2]-HPDI(samples,prob=0.99)[1])
  i_s <- c(i_s,i)
}
data.frame(
  inter,i_s
)

```

## Hard

The Hard problems here all use the data below. These data indicate the gender (male=1, fe- male=0) of officially reported first and second born children in 100 two-child families.

```{r}
library(rethinking)
data(homeworkch3)

```

## 3H1. 

Using grid approximation, compute the posterior distribution for the probability of a birth being a boy. Assume a uniform prior probability. Which parameter value maximizes the posterior probability?

```{r}
p_grid <- seq( from=0 , to=1 , length.out=1000 )
prior <- rep( 5 , 1000 )
likelihood <- dbinom( sum(birth1)+sum(birth2) , 
                      size=length(birth1)+length(birth2) , 
                      prob=p_grid )
posterior <- likelihood * prior
posterior <- posterior / sum(posterior)
print(p_grid[posterior==max(posterior)])

```

## 3H2. 

Using the sample function, draw 10,000 random parameter values from the posterior distribution you calculated above. Use these samples to estimate the 50%, 89%, and 97% highest posterior density intervals.

```{r}
samples <- sample(p_grid, 1e4, replace=T, prob = posterior)
c(HPDI(samples, prob=0.5),HPDI(samples, prob=0.89),HPDI(samples, prob=0.97))
```

## 3H3. 

Use rbinom to simulate 10,000 replicates of 200 births. You should end up with 10,000 numbers, each one a count of boys out of 200 births. Compare the distribution of predicted numbers of boys to the actual count in the data (111 boys out of 200 births). There are many good ways to visualize the simulations, but the dens command (part of the rethinking package) is probably the easiest way in this case. Does it look like the model fits the data well? That is, does the distribution of predictions include the actual observation as a central, likely outcome?

```{r}
stuff=rbinom(n=1e4,size=200,prob=samples)

dens(stuff)
abline( v=sum(birth1)+sum(birth2) , col="red" )
```

## 3H4. 

Now compare 10,000 counts of boys from 100 simulated first borns only to the number of boys in the first births, birth1. How does the model look in this light?

```{r}
likelihood_1 <- dbinom( sum(birth1) , 
                      size=length(birth1) , 
                      prob=p_grid )
posterior_1 <- likelihood_1 * prior
posterior_1 <- posterior_1 / sum(posterior_1)
samples_1 <- sample(p_grid, 1e4, replace=T, prob = posterior_1)
stuff_1=rbinom(n=1e4,size=100,prob=samples_1)
dens(stuff_1)
abline( v=sum(birth1) , col="red" )
```

## 3H5. 

The model assumes that sex of first and second births are independent. To check this assumption, focus now on second births that followed female first borns. Compare 10,000 simulated counts of boys to only those second births that followed girls. To do this correctly, you need to count the number of first borns who were girls and simulate that many births, 10,000 times. Compare the counts of boys in your simulations to the actual observed count of boys following girls. How does the model look in this light? Any guesses what is going on in these data?

```{r}
n_girls1 <- length(birth1)-sum(birth1)
boy_after_girl <- sum(birth2[birth1==0])
stuff_3=rbinom(n=1e4,size=n_girls1,prob=samples_1)
dens(stuff_3)
abline( v=boy_after_girl , col="red" )
```

## Extras

### A2

Suppose the globe tossing data (Chapter 2) had turned out to be 3W and 11L. Construct the posterior distribution. then using the posterior distribution, compute the posterior predictive distribution for the next 5 tosses of the same globe.

```{r}
likelihood_a2 <- dbinom(3  , 
                      size=11 , 
                      prob=p_grid )
posterior_a2 <- likelihood_a2 * prior
posterior_a2 <- posterior_a2 / sum(posterior_a2)
samples_a2 <- sample(p_grid, 1e4, replace=T, prob = posterior_a2)
stuff_a2=rbinom(n=1e4,size=5,prob=samples_a2)
dens(stuff_a2)

```
