---
title: "A06 & 7 Elemental confounds"
author: "David Mendez"
date: "2026-03-02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Book

## 9.1. Goog king Markov and his island kingdom

```{r}
library(rethinking)
num_weeks <- 1e5
positions <- rep(0, num_weeks)
current <- 10
for (i in 1:num_weeks){
        positions[i] <- current
        proposal <- current + sample (c(-1,1),size=1)
        if (proposal < 1) proposal <- 10
        if (proposal > 10) proposal <- 1
        prob_move <- proposal/current
        current <-  ifelse(runif(1)<prob_move, proposal, current)
}
plot(1:100, positions[1:100])
plot(table(positions))
```

## 9.2. Metropolis algorithms

```{r}
D <- 10
T <- 1e3
Y <- rmvnorm(T, rep(0,D),diag(D))
rad_dist <- function(Y) sqrt(sum(Y^2))
Rd <- sapply(1:T, function(i) rad_dist(Y[i,]))
dens(Rd)
```

## 9.4. Easy HMC: ulam

```{r}
library(rethinking)
data("rugged")
d <- rugged
d$log_gdp <- log(d$rgdppc_2000)
dd <- d[complete.cases(d$rgdppc_2000),]
dd$log_gdp_std <- dd$log_gdp / mean(dd$log_gdp)
dd$rugged_std <- dd$rugged / max(dd$rugged)
dd$cid <- ifelse(dd$cont_africa==1,1,2)
```

```{r}
m8.3 <- quap(
        alist(
                log_gdp_std ~dnorm(mu, sigma),
                mu <- a[cid] + b[cid]*(rugged_std - 0.215),
                a[cid] ~ dnorm(1,0.1),
                b[cid] ~dnorm (0, 0.3),
                sigma ~dexp(1)
        ), data=dd
)
precis(m8.3, depth=2)
```

```{r}
dat_slim <- list(
        log_gdp_std = dd$log_gdp_std,
        rugged_std = dd$rugged_std,
        cid = as.integer(dd$cid)
)
m9.1 <- ulam(
        alist(
                log_gdp_std ~ dnorm(mu, sigma),
                mu <- a[cid] + b[cid]*(rugged_std - 0.215),
                a[cid] ~ dnorm(1,0.1),
                b[cid] ~ dnorm(0, 0.3),
                sigma ~dexp(1)
        ), data=dat_slim, chains=1
)
precis(m9.1, depth=2)
```

```{r}
m9.1 <- ulam(
        alist(
                log_gdp_std ~ dnorm(mu, sigma),
                mu <- a[cid] + b[cid]*(rugged_std - 0.215),
                a[cid] ~ dnorm(1,0.1),
                b[cid] ~ dnorm(0, 0.3),
                sigma ~dexp(1)
        ), data=dat_slim, chains=4, cores=4
)
show(m9.1)

```

```{r}
pairs(m9.1)
traceplot(m9.1)
trankplot(m9.1)
```

## 9.5. care and feeding of your Markov Chain

```{r}
y <- c(-1,1)
set.seed(11)
m9.2 <- ulam(
        alist(
                y ~ dnorm( mu, sigma ),
                mu <- alpha,
                alpha ~ dnorm(0, 1e3),
                sigma ~ dexp( 0.0001 )
        ), data = list(y=y), chains=3
)
precis(m9.2)
```

```{r}
traceplot(m9.2)
trankplot(m9.2)
```

```{r}
set.seed(11)
m9.3 <- ulam(
        alist(
                y ~ dnorm( mu, sigma ),
                mu <- alpha,
                alpha ~ dnorm(0, 10),
                sigma ~ dexp( 1 )
        ), data = list(y=y), chains=3
)
precis(m9.3)
traceplot(m9.3)
trankplot(m9.3)
```

### 9.5.4. Non identifiable parameters

```{r}
set.seed(41)
y <- rnorm(100, mean=0, sd=1)
m9.4 <- ulam(
        alist(
                y ~ dnorm( mu, sigma ),
                mu <- a1 + a2,
                a1 ~ dnorm(0, 1e3),
                a2 ~ dnorm(0, 1e3),
                sigma ~ dexp( 1 )
        ), data = list(y=y), chains=3
)
precis(m9.4)
traceplot(m9.4)
trankplot(m9.4)
```

```{r}
m9.5 <- ulam(
        alist(
                y ~ dnorm( mu, sigma ),
                mu <- a1 + a2,
                a1 ~ dnorm(0, 10),
                a2 ~ dnorm(0, 10),
                sigma ~ dexp( 1 )
        ), data = list(y=y), chains=3
)
precis(m9.5)
traceplot(m9.5)
trankplot(m9.5)
```

# Practice

## Easy

### 9E1. v

Which of the following is a requirement of the simple Metropolis algorithm?

1.  The parameters must be discrete.

2.  The likelihood function must be Gaussian.

3.  The proposal distribution must be symmetric.

    Number 3, probability of jumping from A to B must be the same as B to A.

### 9E2. v

Gibbs sampling is more efficient than the Metropolis algorithm. How does it achieve this extra efficiency? Are there any limitations to the Gibbs sampling strategy?

It achieves good performance by using conjugate pairs of priors and likelyhoods since they have an analytical solution for the posterior. It breaks down for high dimensionality problems and nasty priors.

### 9E3. v

Which sort of parameters can Hamiltonian Monte Carlo not handle? Can you explain why?

Continuous parameters since it doesn't jump from sample to sample, it glides from one to the other.

### 9E4. v

Explain the difference between the effective number of samples, n_eff as calculated by Stan, and the actual number of samples.

Stan requires a warm up so during the warm-up, samples are taken but are not used.

### 9E5. v

Which value should Rhat approach, when a chain is sampling the posterior distribution correctly?

1

### 9E6. v

Sketch a good trace plot for a Markov chain, one that is effectively sampling from the posterior distribution. What is good about its shape? Then sketch a trace plot for a malfunctioning Markov chain. What about its shape indicates malfunction?

It's random, stable and symmetric.

### 9E7. v

Repeat the problem above, but now for a trace rank plot.

Same

## Medium

### 9M1.

Re-estimate the terrain ruggedness model from the chapter, but now using a uniform prior for the standard deviation, sigma. The uniform prior should be dunif(0,1). Use ulam to estimate the posterior. Does the different prior have any detectible influence on the posterior distribution of sigma? Why or why not?

```{r}
dat_slim <- list(
        log_gdp_std = dd$log_gdp_std,
        rugged_std = dd$rugged_std,
        cid = as.integer(dd$cid)
)
m9.1 <- ulam(
        alist(
                log_gdp_std ~ dnorm(mu, sigma),
                mu <- a[cid] + b[cid]*(rugged_std - 0.215),
                a[cid] ~ dnorm(1,0.1),
                b[cid] ~ dnorm(0, 0.3),
                sigma ~dexp(1)
        ), data=dat_slim, chains=1
)
m9.3a <- ulam(
        alist(
                log_gdp_std ~ dnorm(mu, sigma),
                mu <- a[cid] + b[cid]*(rugged_std - 0.215),
                a[cid] ~ dnorm(1,0.1),
                b[cid] ~ dnorm(0, 0.3),
                sigma ~ dunif(0,1)
        ), data=dat_slim, chains=1
)
precis(m9.1, depth=2)
precis(m9.3a, depth=2)
```

Not really, likely because both priors are very uninformative.

### 9M2.

Modify the terrain ruggedness model again. This time, change the prior for b[cid] to dexp(0.3). What does this do to the posterior distribution? Can you explain it?

```{r}
m9.4a <- ulam(
        alist(
                log_gdp_std ~ dnorm(mu, sigma),
                mu <- a[cid] + b[cid]*(rugged_std - 0.215),
                a[cid] ~ dnorm(1,0.1),
                b[cid] ~ dexp(0.3),
                sigma ~ dexp(1)
        ), data=dat_slim, chains=1
)
precis(m9.4a, depth = 2)
```

It flips the sign of one of the b parameters since the exponential distirbution is always positive. The rest of the parameters remain positive.

### 9M3.

Re-estimate one of the Stan models from the chapter, but at different numbers of warmup iterations. Be sure to use the same number of sampling iterations in each case. Compare the n_eff values. How much warm-up is enough?

## Hard

## Extra
